Data Management \\ and LSST Special Programs

### 2017-06-09 ms.tex
organizational edits

{\it Mario: "Examples include making sure there's a requirement that the system is able to use an arbitrary template for some (deep drilling) field and a plan to deliver that functionality; or that the requirements define how a user-supplied code will be able to run in the batch system to process the last-night's data with shift-and-stack code for KBOs."}

% % % % % % % % % % % % % % % % % % 
\subsection{Boundaries of Allowed Special Programs Observations}\label{ssec:intro_bounds}

In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community. The LSST project will need to define boundaries for these observations. Although not all of these boundaries will originate with DM constraints, we start that process here (\textcolor{red}{MLG: Are any initially derived boundaries for Special Programs observations specified elsewhere?}). The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}

Start of a list of observing restrictions. \\
\textbf{Sky coordinates --} (The RA and Dec that will be accessible from the site.) \\
\textbf{Pointing alt/az, airmass --} (Telescope pointing constraints.) \\
\textbf{Sky background level --} (e.g., twilight observations cannot start before X) \\
\textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
\textbf{Filters --} (e.g., list filter combinations; $gri$ always be on?; change times, etc.) \\
\textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD Main Survey). \\

@@

To resolve some open issues we will need to solicit input from the science collaborations (SC). However, it may be better and easier to wait until the next call for community input, and package these requests up in a broader request that a kind of "DM Requirements" section be included in all white papers. The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018.

\noindent Recall that there is already a community forum on this topic, so far minimally used but ready to be revised in the future when the next calls soliciting white papers are released. \\ \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.

\noindent Remember also that the SC Coordinator for DDF: Neil Brandt.

@@

% % % % % % % % % % % % % % % % % % 
\subsection{A Summary of Open Action Items} \label{ssec:intro_AI}

\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications. \end{enumerate}
\begin{enumerate}[topsep=-15pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
\begin{enumerate}[resume,topsep=-15pt,label= \textbf{Action \Roman*.}] \item Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the lower and upper limits on exposure time. \end{enumerate}

@@

\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.}] \item Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}

@@

\noindent \textbf{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\\
The OSS contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from the SRD and SR). There are two points worth mentioning: \\
$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
$\rightarrow$ There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times. However, there doesn't seem to be a specified limit on 'thermal stability'. Knowing the full range of exposure times and image cadences that DM must consider is important for estimating the predicted latencies providing raw, reduced and calibrated images (Section \ref{ssec:dmdocs_L3}) and testing the incorporation of non-standard images into the calibration and difference imaging pipelines (Section \ref{ssec:dmdocs_SPinWFD_L1}).
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications. \end{enumerate}
$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
$\rightarrow$ The specified maximum filter change time could be a debilitating overhead for, e.g., a deep drilling cadence of consecutive 30 second images wants to rotate through filters. Furthermore, we might expect that in reality the filter change time will be shorter than 120 seconds, which would enable a DDF to sequence filters. Furthermore, while the minimum filter change count is not an issue, I can't find the nightly (or total lifetime) maximum to limit wear and tear on the mechanism (as I've heard about). However, there is already a change request in progress regarding this issue: \url{https://project.lsst.org/groups/ccb/node/938}.

\noindent \textbf{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}} \\
The DMSR contains the high-level requirements for all processing and products. There are two main areas that pertain to Special Programs. \\
$\bullet$ Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069)\\
$\rightarrow$ Specifies that {\it "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing. They will be re-generated for users on-demand using the latest processing software and calibrations"}, but also says that {\it "This aspect of the processing for Special Programs data is specific to each program"}. It is unclear whether processed SP data will remain on disk and accessible for as long as it is required by the science users, or if it will only live a short while.
\begin{enumerate}[topsep=-10pt,label= \textbf{Action \Roman*.},resume] \item Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request. \end{enumerate}
$\bullet$ Section 2.9, "Level 3 Production" \\
Specifies services of the DMS for Level 3 processing which are relevant to Special Programs: \\
2.9.1. Level 3 Data Import (ID: DMS-REQ-0290): to provide the ability to ingest common file formats (e.g., FITS table of external catalog). \\
2.9.2. Resource Allocation (ID: DMS-REQ-0119): to provide a mechanism to prioritize Level 3 activities and appropriately allocate processing resources. \\
2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. \\
2.9.4. Provenance (ID: DMS-REQ-0121): to provide a means for recording information about processing performed at DACs (DMS-provided input and user inputs) to help users work in a reproducible way. \\
2.9.5/6. Software Frameworks (ID: DMS-REQ-0125,DMS-REQ-0128): to provide a means for applying user-provided processing to catalog and image data, and for assessing the completeness of that application. \\
3.1. Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops. \\
$\rightarrow$ \textcolor{red}{Questions}: Regarding 2.9.3, I'm not sure what {\it "self-consistent inputs"} means and so cannot assess any potential issues. Regarding 3.1, I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?

@@

For each of these questions or potential issues, a response will be: \\
\textcolor{blue}{blue if the answer is known} (even if the answer is a negative), \\
\textcolor{red}{red if the answer is still to be determined}. \\
If it seems particularly urgent or involved to answer, it will spawn an Action Item.

@@

$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1, the higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA. Is there a science need for 60 second latency on a consecutive time-series of short exposures, or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred? (I.e., the main goal of shorter exposures would be to include brighter stars, not necessarily to monitor variability on short timescales). Alternatively, are there options for short-term increases in parallel processing power at NCSA that we could invoke during times of high cadence observations? \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Query the science community and NCSA about the science needs and processing capabilities, respectively, of $<30$ second cadence imaging. \end{enumerate}

$\bullet$ LDM-151, Section 3.3.4 `Alert queuing and persistance' mentions that the {\it ``event message stream and the AlertDB will be synchronized at least once every 24 hours"} and that {\it ``Prior to the start of the subsequent night's observations, the message queue will be flushed and synchronized with the AlertDB. It is possible to persist the message queue on longer timescale but it is a requirement that synchronization be performed within 24 hours of the observations."} Since this is for the main survey in which new alerts are issued for a given field 2--3 times per night, check that this is fine in the case of doing e.g., 200 alert bursts on the same field in a single night from a deep drilling field? \\
$\rightarrow$  \textcolor{red}{Unanswered.} However, upon review, this requirement seems unrelated to visit locations, and thus not an issue for Special Programs?

@@

$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used, but this will be needed for both Levels 1 and 3 differencing pipelines and products. \\
$\rightarrow$ \textcolor{red}{Unaddressed.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}

$\bullet$ LDM-151 mentions that forced photometry will be done for externally defined targets (Section 3.2.5), and this may be a particular interest to Special Programs, but it is unclear how such targets will be identified or flagged as such in the databases, and unclear whether we need to add an element to the database schema for this. Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType. Indicates which input was used to produce a given object."} Is that all we need? \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. \end{enumerate}

$\bullet$ Will the following database element sizes be enough to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)? Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program, we need to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales. \\
{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on whether {\tt float[6 x 32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}

$\bullet$ Low surface-brightness features and/or faint extended objects are required for some galaxy science, and will show up more in e.g., a DDF than a WFD CoAdd. Will the related shape parameters such as extendedness in the {\tt Object} database be sufficient to describe the extra level of detail in low-surface brightness features that might come from deeper Special Programs CoAdds? As above, these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program. Furthermore, it's unlikely that extremely deep patches will be included in the Level 2 products, but this still should be assessed.\\
$\rightarrow$ \textcolor{red}{Unassessed.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Action \Roman*.}] \item Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}

@@

$\bullet$ LDM-151 says (Section 3.2.4.1) that the image differencing pipeline will measure the PSF flux on snap difference images for all {\tt DIASources}. Could Special Programs data that is only a single exposure, not a visit that is the combination of two snaps, be incorporated into the Level 1 AP? \\
$\rightarrow$ \textcolor{red}{Unanswered.}

@@

The following is a collection of rough notes on the science goals for some proposed Special Programs, classified by area of astronomy. It is not comprehensive. Any major questions that were inspired by compiling these science goals have been incorporated into Section \ref{sec:issues}, so no Action Items are spawned in this section (although there are some low-level unanswered questions).

>>>>

{\it Current Status: This document is a rough draft, with the goal of being exhaustively complete. Many of the identified \textbf{``Concerns"} may just be misunderstandings or a lack of knowledge.}

The rest of this introduction contains: \\
\S~\ref{ssec:intro_bounds} -- defining DM-related boundaries for Special Programs. \\
\S~\ref{ssec:intro_forum} -- existing channels for interaction with Science communities. \\
\S~\ref{ssec:intro_UC} -- a summary of unresolved \textbf{Concerns} that arose in this study.

Later sections attempt to identify potential \textbf{Concerns} regarding DM's plans for processing the data from Special Programs with a variety of approaches: \\
\S~\ref{sec:dmdocs} -- a review of the current DM plans for SP data that are either already explicitly written down in DM documentation, or exist in Mario's thoughts. \\
\S~\ref{sec:issues} -- a brainstorm of potential issues that might arise, organized by pipeline or product, inspired by internal and external documentation. \\
\S~\ref{sec:science} -- a review of science cases for SP and their potential strain on the DM system, including detailed reduction/analysis case studies in \S~\ref{ssec:science_dmsums}.

\clearpage
% % % % % % % % % % % % % % % % % % 
\subsection{Boundaries of Allowed Special Programs Observations}\label{ssec:intro_bounds}

In the next couple of years, cadences and pointings for DDF and new mini-surveys will be proposed by the community. The document "Cadence Exploration and Improvement Plan" (by Zeljko, Lynne, Andy) proposes that the next call for DDF White Papers (proposed fields and cadences) be in December 2017, and the call for Mini-Survey White Papers be in October 2018. We will need to define boundaries for these observations from a DM prospective. The relevant Jira issue is \url{https://jira.lsstcorp.org/browse/LIT-325?filter=-3}.

Start of a list of DM-related considerations for Special Programs data. \\
\textbf{Exposure times --} (e.g., current shortest is 1 s, with stretch goal of 0.1, but... \\
\textbf{Afternoon calibrations --} (e.g., must fit into some amount of time left over after the WFD main survey daily calibration frames are obtained) \\
\textbf{Time allotted --} Nominally, 10\% of the total observing time (90\% goes to the WFD).

@@

To resolve some open issues we will need to solicit input from the science collaborations (SC; also, remember that the SC Coordinator for DDF is Neil Brandt). It may be better and easier to wait until the next call for community input, and request that a "DM Requirements" section be included in all white papers (as in \S~\ref{ssec:science_dmsums}). Also, there is already a community forum on DDF, so far minimally used but ready to be revived when the next calls soliciting white papers are released. \url{http://community.lsst.org/t/deep-drilling-fields-and-data-management/1115}.

@@

\clearpage
% % % % % % % % % % % % % % % % % % 
\subsection{A Summary of Unresolved Concerns} \label{ssec:intro_UC}

\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications. \end{enumerate}
\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C12} Confirm whether DM's ISR algorithms will be developed to reduce images with a range of exposure times, dither patterns, etc., and if not, what are the data restrictions for ISR. \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C13} Will there be a Science Platform capability set up so that users are notified not just when any data for their Special Program is taken, but when certain sets are finished and ready for processing? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C14} Does the Level 3 pipeline have to include the steps to regenerate the processed exposures, or will DM deliver a service to automatically initiate regeneration by e.g., simply calling for an exposure or difference image? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C15} Is it yet known how users will apply for batch processing priority and how they will be notified their jobs have finished, e.g., RSS feed? I could imagine that this is a requirement that would be defined in a future document on the Science Platform deliverables? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C16} Would the internal real/bogus routine be able to run on nightly CoAdd DDF difference images? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C17} How long can some Level 3 products remain on disk, such as template images for nightly CoAdd image subtraction? \end{enumerate}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C18} Confirm that the idea of federated Level 3 codes still exists (e.g., given how the conversation on incorporating a photo-$z$ code ended up in a proposed descope). \end{enumerate}

@@

\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C1} Investigate the minimum and maximum exposure times that can be successfully used in difference imaging and thereby contribute to the Alert Stream. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C2} Confirm whether a latency shorter than {\tt \textbf{L1PublicT}}=24 hours for Level 1 data products (LSR-REQ-0104) can be provided to federated and/or user-run Level 3 pipelines for Special Programs. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C3} Specify the latencies for delivering raw, reduced, and calibrated Special Programs images that are not processed by the Level 1 pipeline. \end{enumerate}

@@

\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C4} Confirm whether the computational budget for processing the Special Programs data requires more than 10\% of the total resources, and on which batch system the bulk of this processing will take place. \end{enumerate}

@@

\noindent \textbf{Observatory System Specifications, OSS, LSE-30, \cite{LSE-30}}\\
The OSS contains high-level requirements for the Level 3 data products, and their storage and access, that are similar to those in the DMSR (as they both stem from the SRD and SR). There are two points worth mentioning: \\
$\bullet$ 3.6.1.4. Minimum exposure time (ID: OSS-REQ-0291): 1 second (stretch 0.1 seconds). \\
$\rightarrow$ There is a note that if the exposure time is shortened from 15 seconds, the spacing between exposures may need to be lengthened in order to maintain camera thermal stability, and that the thermal stability might also be affected with longer exposure times. However, there doesn't seem to be a specified limit on 'thermal stability'. Knowing the full range of exposure times and image cadences that DM must consider is important for estimating the predicted latencies providing raw, reduced and calibrated images (Section \ref{ssec:dmdocs_L3}) and testing the incorporation of non-standard images into the calibration and difference imaging pipelines (Section \ref{ssec:dmdocs_SPinWFD_L1}). The science community will also need this information when they propose their DDF cadences and write their Special Programs white papers.
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C5} Confirm the expected thermal tolerances of the CCD and the associated minimum and maximum exposure times and exposure cadence that will keep the camera temperature within specifications. \end{enumerate}
$\bullet$ 3.6.2.2. Maximum time for operational filter change (ID: OSS-REQ-0293): 120 seconds \\
$\bullet$ 3.6.2.4. Minimum filter change count (ID: OSS-REQ-0295): day, 8, night, 4 \\
$\rightarrow$ The specified maximum filter change time could be a debilitating overhead for, e.g., a deep drilling cadence of consecutive 30 second images wants to rotate through filters. Furthermore, we might expect that in reality the filter change time will be shorter than 120 seconds, which would enable a DDF to sequence filters. Furthermore, while the minimum filter change count is not an issue, I can't find the nightly (or total lifetime) maximum to limit wear and tear on the mechanism (as I've heard about). However, there is already a change request in progress regarding this issue: \url{https://project.lsst.org/groups/ccb/node/938}.

\noindent \textbf{Data Management Subsystems Requirements, DMSR, LSE-61, \cite{LSE-61}} \\
The DMSR contains the high-level requirements for all processing and products. There are two main areas that pertain to Special Programs. \\
$\bullet$ Section 1.2.2 "Processed Visit Images" (ID: DMS-REQ-0069)\\
$\rightarrow$ Specifies that {\it "Processed science exposures are not archived, and are retained for only a limited time to facilitate down-stream processing. They will be re-generated for users on-demand using the latest processing software and calibrations"}, but also says that {\it "This aspect of the processing for Special Programs data is specific to each program"}. It is unclear whether processed SP data will remain on disk and accessible for as long as it is required by the science users, or if it will only live a short while.
\begin{enumerate}[topsep=-10pt,label= \textbf{Concern \Roman*.},resume] \item \label{C6} Confirm whether processed single visits for Special Programs can stay on disk and accessible for Level 3 pipelines for a longer amount of time, by user request. \end{enumerate}
$\bullet$ Section 2.9, "Level 3 Production" \\
Specifies services of the DMS for Level 3 processing which are relevant to Special Programs: \\
2.9.1. Level 3 Data Import (ID: DMS-REQ-0290): to provide the ability to ingest common file formats (e.g., FITS table of external catalog). \\
2.9.2. Resource Allocation (ID: DMS-REQ-0119): to provide a mechanism to prioritize Level 3 activities and appropriately allocate processing resources. \\
2.9.3. Data Product Self-Consistency (ID: DMS-REQ-0120): to provide a means for ensuring Level 3 tasks can be carried out on self-consistent inputs. \\
2.9.4. Provenance (ID: DMS-REQ-0121): to provide a means for recording information about processing performed at DACs (DMS-provided input and user inputs) to help users work in a reproducible way. \\
2.9.5/6. Software Frameworks (ID: DMS-REQ-0125,DMS-REQ-0128): to provide a means for applying user-provided processing to catalog and image data, and for assessing the completeness of that application. \\
3.1. Software for Community Re-Use (ID: DMS-REQ-0308): algorithms shall be designed to run on both high-performance platforms and users' desktops. \\
$\rightarrow$ \textcolor{red}{Questions}: Regarding 2.9.3, I'm not sure what {\it "self-consistent inputs"} means and so cannot assess any potential issues. Regarding 3.1, I thought that the Science Platform was being developed specifically so that no one needs to process LSST data on their home machines?

@@

For each of these questions or potential issues, a response will be: \\
\textcolor{blue}{blue if the answer is known} (even if the answer is a negative), \\
\textcolor{red}{red if the answer is still to be determined}. \\
If it seems particularly urgent or involved to answer, it will spawn a \textbf{Concern}.

@@

$\bullet$ When we attempt to include short exposures (e.g., 5 seconds) in Level 1, the higher data acquisition rate (e.g., $\sim$6$\times$) might cause a processing back-up at NCSA. Is there a science need for 60 second latency on a consecutive time-series of short exposures, or would e.g., 60 seconds latency on every 6$^{\rm th}$ image, be preferred? (I.e., the main goal of shorter exposures would be to include brighter stars, not necessarily to monitor variability on short timescales). Alternatively, are there options for short-term increases in parallel processing power at NCSA that we could invoke during times of high cadence observations? \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C7} Query the science community about their science needs, and NCSA about their processing capabilities, for $<30$ second cadence imaging. \end{enumerate}

$\bullet$ LDM-151, Section 3.3.4 `Alert queuing and persistance' mentions that the {\it ``event message stream and the AlertDB will be synchronized at least once every 24 hours"} and that {\it ``Prior to the start of the subsequent night's observations, the message queue will be flushed and synchronized with the AlertDB. It is possible to persist the message queue on longer timescale but it is a requirement that synchronization be performed within 24 hours of the observations."} Since this is for the main survey in which new alerts are issued for a given field 2--3 times per night, check that this is fine in the case of doing e.g., 200 alert bursts on the same field in a single night from a deep drilling field? \\
$\rightarrow$  \textcolor{red}{Unanswered.} However, upon review, this requirement seems unrelated to visit locations, and thus not an issue for Special Programs.

@@

$\bullet$ The database schema for {\tt DIASource} does not appear to have an element that identifies which template image that was used, but this will be needed for both Levels 1 and 3 differencing pipelines and products. \\
$\rightarrow$ \textcolor{red}{Unaddressed.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C8} Database schema for {\tt DIASource} needs to have an element added that contains information about the template that was used to create the difference image. \end{enumerate}

$\bullet$ LDM-151 mentions that forced photometry will be done for externally defined targets (Section 3.2.5), and this may be a particular interest to Special Programs, but it is unclear how such targets will be identified or flagged as such in the databases, and unclear whether we need to add an element to the database schema for this. Currently, the {\tt Object} database contains an element {\tt prv\_inputId} which is an {\tt integer}, and is described as the {\it ``Pointer to prv\_InputType. Indicates which input was used to produce a given object."} Is that all we need? \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C9} Clarify whether {\tt Object.prv\_inputID} will identify whether an object is an externally provided coordinate for forced photometry. \end{enumerate}

$\bullet$ Will the following database element sizes be enough to characterize the different kinds of variability that could be measured with Special Programs cadences that are quite different from the main survey (e.g., Miras, flares, and LBV-turned-SNe)? Although these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program, we need to ensure that they are adequate for the Level 1 database, which might incorporate observations from Special Programs on a variety of timescales. \\
{\tt Object} and {\tt DIAObject.lcPeriodic} = {\tt float[6 x 32]} = Periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
{\tt Object} and {\tt DIAObject.lcNonPeriodic} = {\tt float[6 x 32]} = Non-periodic features extracted from light-curves using generalized Lomb-Scargle periodogram \\
$\rightarrow$ \textcolor{red}{Unanswered.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C10} Contact the Science Collaborations for input on whether {\tt float[6x32]} for the characterization parameters for variability is adequate to express the additional timescales and depths covered by all possible Special Programs observing strategies. \end{enumerate}

$\bullet$ Low surface-brightness features and/or faint extended objects are required for some galaxy science, and will show up more in e.g., a DDF than a WFD CoAdd. Will the related shape parameters such as extendedness in the {\tt Object} database be sufficient to describe the extra level of detail in low-surface brightness features that might come from deeper Special Programs CoAdds? As above, these elements could be differently populated in the Level 3 databases in order to meet the science goals of a given Special Program. Furthermore, it's unlikely that extremely deep patches will be included in the Level 2 products, but this still should be assessed.\\
$\rightarrow$ \textcolor{red}{Unassessed.}
\begin{enumerate}[resume,topsep=-10pt,label= \textbf{Concern \Roman*.}] \item \label{C11} Contact the Science Collaborations for input on the characterization parameters for extended objects. \end{enumerate}

@@

$\bullet$ LDM-151 says (Section 3.2.4.1) that the image differencing pipeline will measure the PSF flux on snap difference images for all {\tt DIASources}. Could Special Programs data that is only a single exposure, not a visit that is the combination of two snaps, be incorporated into the Level 1 AP? \\
$\rightarrow$ \textcolor{red}{Unanswered.} But since there is talk anyway of using 1x30 instead of 2x15 snaps, I am not spawning a \textbf{Concern} for this question.

@@

The following is a collection of rough notes on the science goals for some proposed Special Programs, classified by area of astronomy. It is not comprehensive. Any major questions that were inspired by compiling these science goals have been incorporated into Section \ref{sec:issues}, so no \textbf{Concerns} are spawned in this section (although there are some low-level unanswered questions).

\medskip
\noindent \textbf{Solar System Objects}\\
$\bullet$ Observations in the North Ecliptic Spur region will yield more $\geq140$ m NEOs (Brandt talk). \\
$\bullet$ Detecting faint SSO down to $r\sim27$ by applying shift-and-stack processing to a mini-survey (Trans-Neptunian Objectss, Trojans, asteroids, long-period comets, dwarf planets). \citep{BeckerWP} \\
$\rightarrow$ Regarding shift-and-stack, \cite{BeckerWP} says that ``the multi-fit algorithm ... naturally provides a base infrastructure for this process. In particular, the marshaling of the pixels to attempt a given photometric measurement is non-trivial when tens of thousands of images are required. However, the multi-fit middleware is required to do exactly this, so we expect that this issue will be resolved by the time SAS is needed." Is this true? \textcolor{red}{Unanswered}, but this does not seem like an urgent concern for DM, so no \textbf{Concern} is spawned. \\
$\bullet$ Mini-moons, temporarily captured satellites of the Earth. (Section 10.2, \cite{OSWP}) \\
$\bullet$ Find meter-sized impactors $<2$ weeks before impact.  (Section 10.2, \cite{OSWP}) \\
$\bullet$ Sweetspot: twilight fields to find NEOs in Earth-like orbits (never in opposition fields). (Section 10.2, \cite{OSWP})
